---
title: "Classifying Effects Using Confidence Limits"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This tutorial demonstrates how to use the `add_effect_classification()` function to classify effect sizes based on confidence intervals. This is useful when you want to interpret uncertainty in trends or treatment effects using a transparent and rule-based approach.

## About effect classification

The `add_effect_classification()` function takes a dataframe with confidence limits and adds classification columns that describe:

- the **direction** and **strength** of the effect (e.g., *strong increase*, *moderate decrease*, *unknown*)
- both **fine-grained** and **coarse** interpretations.

It uses the [**effectclass**](https://inbo.github.io/effectclass/) package internally to perform the classification logic.
Classification is based on:

1. the lower and upper **confidence limits** around an effect estimate
2. a **reference value** (e.g., 0 for no effect)
3. and a **threshold** or range of thresholds that define what counts as a meaningful effect

Thresholds and reference values should reflect scientifically or practically meaningful changes — e.g., a biologically relevant increase or decline in a biodiversity indicator.

See the [effectclass tutorial](https://inbo.github.io/effectclass/articles/classification.html) for a more elaborate overview.

<!-- spell-check: ignore:start -->
<table style="table-layout: auto; width: 100%;">
  <thead>
    <tr>
      <th style="white-space: nowrap;">Symbol</th>
      <th style="white-space: nowrap;">Fine effect / trend</th>
      <th style="white-space: nowrap;">Coarse effect / trend</th>
      <th style="white-space: nowrap;">Rule</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td><code>++</code></td>
      <td>strong positive effect / strong increase</td>
      <td>positive effect / increase</td>
      <td>confidence interval above the upper threshold</td>
    </tr>
    <tr>
      <td><code>+</code></td>
      <td>positive effect / increase</td>
      <td>positive effect / increase</td>
      <td>confidence interval above reference and contains the upper threshold</td>
    </tr>
    <tr>
      <td><code>+~</code></td>
      <td>moderate positive effect / moderate increase</td>
      <td>positive effect / increase</td>
      <td>confidence interval between reference and the upper threshold</td>
    </tr>
    <tr>
      <td><code>~</code></td>
      <td>no effect / stable</td>
      <td>no effect / stable</td>
      <td>confidence interval between thresholds and contains reference</td>
    </tr>
    <tr>
      <td><code>-~</code></td>
      <td>moderate negative effect / moderate decrease</td>
      <td>negative effect / decrease</td>
      <td>confidence interval between reference and the lower threshold</td>
    </tr>
    <tr>
      <td><code>-</code></td>
      <td>negative effect / decrease</td>
      <td>negative effect / decrease</td>
      <td>confidence interval below reference and contains the lower threshold</td>
    </tr>
    <tr>
      <td><code>--</code></td>
      <td>strong negative effect / strong decrease</td>
      <td>negative effect / decrease</td>
      <td>confidence interval below the lower threshold</td>
    </tr>
    <tr>
      <td><code>?+</code></td>
      <td>potential positive effect / potential increase</td>
      <td>unknown effect / unknown</td>
      <td>confidence interval contains reference and the upper threshold</td>
    </tr>
    <tr>
      <td><code>?-</code></td>
      <td>potential negative effect / potential decrease</td>
      <td>unknown effect / unknown</td>
      <td>confidence interval contains reference and the lower threshold</td>
    </tr>
    <tr>
      <td><code>?</code></td>
      <td>unknown effect / unknown</td>
      <td>unknown effect / unknown</td>
      <td>confidence interval contains the lower and upper threshold</td>
    </tr>
  </tbody>
</table>
<!-- spell-check: ignore:end -->

## Getting started with dubicube

```{r}
# Load packages
library(dubicube)

# General
library(ggplot2)      # Data visualisation
```

Your dataframe must contain **lower and upper confidence limits**, e.g. `"lcl"` and `"ucl"`.
Let’s start with a synthetic dataset of 10 observations.

```{r}
# Simulated means and standard deviations
ds <- data.frame(
  mean = c(0, 0.5, -0.5, 1, -1, 1.5, -1.5, 0.5, -0.5, 0),
  sd = c(1, 0.5, 0.5, 0.5, 0.5, 0.25, 0.25, 0.25, 0.25, 0.5)
)

# Compute 90% confidence intervals
ds$lcl <- qnorm(0.05, ds$mean, ds$sd)
ds$ucl <- qnorm(0.95, ds$mean, ds$sd)

# View the dataset
ds
```

### Add effect classifications

We use the `add_effect_classification()` for effect classification. It relies on the following arguments:

- **`df`**:
  The input data, a dataframe containing confidence limits, e.g. the result of [bootstrap confidence interval calculation](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-interval-calculation.html).

- **`cl_columns`**:
  A character vector of length 2 specifying the column names in `df` that contain the **lower** and **upper** confidence limits. For example: `cl_columns = c("lcl", "ucl")`.

- **`threshold`**:
  A numeric vector defining the effect size threshold(s) used to distinguish between "no effect" and meaningful increases or decreases. This can be either one number (interpreted symmetrically around the reference) or two (explicit lower and upper thresholds).

- **`reference`**:
  A single numeric value representing the **null hypothesis** or **no-effect** level (e.g. 0). The position of the confidence interval relative to this value determines the direction of the effect.

- **`coarse`**:
  A logical flag indicating whether a simplified classification should also be added. If `TRUE` (default), additional columns `effect_code_coarse` and `effect_coarse` are included, summarizing effects into broader categories (increase, decrease, stable, unknown).


Let’s classify the effects using a threshold of `1` (will be expanded to `[-1, 1]` around the reference) and a reference of `0`.

```{r}
# Perform effect classification
result <- add_effect_classification(
  df = ds,
  cl_columns = c("lcl", "ucl"),
  threshold = 1,
  reference = 0,
  coarse = TRUE
)

# View the result
result
```

### Visualising the result

Detailed guidance on best practices for visualising effect classifications is provided in the tutorials for [temporal trends](https://b-cubed-eu.github.io/dubicube/articles/visualising-temporal-trends.html) and [spatial trends](https://b-cubed-eu.github.io/dubicube/articles/visualising-spatial-trends.html).
Below, we demonstrate a simple and direct way to visualise classified effects using **ggplot2**.

```{r}
# Define coarse colour palette
coarse_colour <- scale_colour_manual(
  values =  c(
    "chartreuse3",
    "gold",
    "firebrick1",
    "skyblue"
  ),
  drop = FALSE
)

# Define fine colour palette
fine_colour <- scale_colour_manual(
  values =  c(
    "darkgreen",
    "chartreuse3",
    "darkolivegreen1",
    "gold",
    "orange",
    "firebrick1",
    "darkred",
    "gray80",
    "gray30",
    "grey55"
  ),
  drop = FALSE
)
```

The following plot shows point estimates and their 90% confidence intervals, coloured by the **coarse** effect classification.

```{r}
#| fig.alt: >
#|   Visualising coarse effect classification.
ggplot(data = result, aes(x = rownames(result))) +
  geom_hline(yintercept = 0, linetype = "longdash", colour = "black") +
  geom_hline(yintercept = c(-1, 1), linetype = "dotdash") +
  geom_errorbar(aes(ymin = lcl, ymax = ucl, colour = effect_coarse),
                linewidth = 1.5, show.legend = TRUE) +
  geom_point(aes(y = mean), colour = "black", size = 3.5) +
  labs(x = "Observation", y = "Effect estimate", colour = "Effect (coarse)") +
  coarse_colour +
  theme_minimal()
```

The plot below provides a more detailed view using the **fine-grained** classification.

```{r}
#| fig.alt: >
#|   Visualising fine effect classification.
ggplot(data = result, aes(x = rownames(result))) +
  geom_hline(yintercept = 0, linetype = "longdash", colour = "black") +
  geom_hline(yintercept = c(-1, 1), linetype = "dotdash") +
  geom_errorbar(aes(ymin = lcl, ymax = ucl, colour = effect),
                linewidth = 1.5, show.legend = TRUE) +
  geom_point(aes(y = mean), colour = "black", size = 3.5) +
  labs(x = "Observation", y = "Effect estimate", colour = "Effect (fine)") +
  fine_colour +
  theme_minimal()
```
