---
title: "Group-Level Sensitivity Analysis"
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

When working with biodiversity data cubes, understanding how sensitive your indicators are to individual groups (like species, sites, or time periods) can be crucial for robust analysis.
This tutorial introduces the `cross_validate_cube()` function from **dubicube**, which helps assess how individual categories influence biodiversity indicators using cross-validation.

## Group-level cross-validation for data cubes

Cross-validation is a resampling technique commonly used in statistics and machine learning to evaluate model performance.
In the context of biodiversity data cubes, we adapt cross-validation to assess how robust indicators are to the inclusion or exclusion of specific groups.
This helps identify influential categories that disproportionately affect indicator values.

### A data cube and a statistic

Consider a data cube $\mathbf{X}$ from which we want to calculate a statistic $\theta$.
The data cube can be grouped, e.g. by `taxon`, that contains multiple categories, e.g. `species1`, `species2`, `species3` ...

-   **Original Sample Data**: $\mathbf{X} = \{X_{11}, X_{12}, X_{13}, \ldots, X_{sn}\}$
    -   The initial set of data points, where there are $s$ different categories in a group (e.g. $s = 10$ species in group `taxon`) and $n$ total samples across all categories (= the sample size). $n$ corresponds to the number of cells in a data cube or the number of rows in tabular format.
-   **Statistic of Interest**: $\theta$
    -   The parameter or statistic being estimated, such as the mean $\bar{X}$, variance $\sigma^2$, or a biodiversity indicator. Let $\hat{\theta}$ denote the estimated value of $\theta$ calculated from the complete dataset $\mathbf{X}$.

### Resampling and recalculating

From $\mathbf{X}$, multiple data cubes $\mathbf{X}_{-s_j}$ are created where in each new dataset a different category is removed compared to the original data cube $\mathbf{X}$.
For example in dataset 1, `species1` is excluded; in dataset 2 `species2`; in dataset 3 `species3`; and so on.
For each new dataset, the statistic $\theta$ is again calculated: $\hat{\theta}_{-s_j}$.

-   **Cross-Validation (CV) Sample**: $\mathbf{X}_{-s_j}$
    -   The full dataset $\mathbf{X}$ excluding all samples belonging to category $j$. This subset is used to investigate the influence of category $j$ on the estimated statistic $\hat{\theta}$.
-   **CV Estimate for Category** $\mathbf{j}$: $\hat{\theta}_{-s_j}$
    -   The value of the statistic of interest calculated from $\mathbf{X}_{-s_j}$, which excludes category $j$. For example, if $\theta$ is the sample mean, $\hat{\theta}_{-s_j} = \bar{X}_{-s_j}$.

### Derivation of error measures

From this, we can calculate different error measures that help inform further analyses or decision-making steps.

-   **Error Measures**:

    -   The **Error** is the difference between the statistic estimated without category $j$ ($\hat{\theta}_{-s_j}$) and the statistic calculated on the complete dataset ($\hat{\theta}$).

    $$
    \text{Error}_{s_j} = \hat{\theta}_{-s_j} - \hat{\theta}
    $$

    -   The **Relative Error** is the absolute error, normalized by the true estimate $\hat{\theta}$ and a small error term $\epsilon = 10^{-8}$ to avoid division by zero.

    $$
    \text{Rel. Error}_{s_j} = \frac{|\hat{\theta}_{-s_j} - \hat{\theta}|}{\hat{\theta} +\epsilon}
    $$

    -   The **Percent Error** is the relative error expressed as a percentage.

    $$
    \text{Perc. Error}_{s_j} = \text{Rel. Error}_{s_j} \times 100 \%
    $$

-   **Summary Measures**:

    -   The **Mean Relative Error (MRE)** is the average of the relative errors over all categories.

    $$
    \text{MRE} = \frac{1}{s} \sum_{j=1}^s \text{Rel. Error}_{s_j}
    $$

    -   The **Mean Squared Error (MSE)** is the average of the squared errors.

    $$
    \text{MSE} = \frac{1}{s} \sum_{j=1}^s (\text{Error}_{s_j})^2
    $$

    -   The **Root Mean Squared Error (RMSE)** is the square root of the MSE.

    $$
    \text{RMSE} = \sqrt{\text{MSE}}
    $$

## Getting started with dubicube

The **dubicube** cross-validation method can be applied on any dataframe from which a statistic is calculated and for which a grouping variable is present.
For this tutorial, we focus on occurrence cubes.
Therefore, we will use the **b3gbi** package for processing the raw data before moving on to the cross-validation.

```{r, message=FALSE, warning=FALSE}
# Load packages
library(ggplot2)      # Data visualisation
library(dplyr)        # Data wrangling

# Data loading and processing
library(frictionless) # Load example datasets
library(b3gbi)        # Process occurrence cubes
library(dubicube)     # Analysis of data quality & reliability
```

### Loading and processing the data

We load the bird cube data from the **b3data** data package using **frictionless** (see also [here](https://github.com/b-cubed-eu/b3data-scripts)).
It is an occurrence cube for birds in Belgium between 2000 en 2024 using the MGRS grid at 10 km scale.

```{r}
# Read data package
b3data_package <- read_package(
  "https://zenodo.org/records/15211029/files/datapackage.json"
)

# Load bird cube data
bird_cube_belgium <- read_resource(b3data_package, "bird_cube_belgium_mgrs10")
head(bird_cube_belgium)
```

We process the cube with **b3gbi** to put the data in a format, ready for analysis.
We select the data from 2011 - 2020.

```{r}
processed_cube <- process_cube(
  bird_cube_belgium,
  first_year = 2011,
  last_year = 2020,
  cols_occurrences = "n"
)
processed_cube
```

### Analysis of the data

Suppose we are interested in the average number of observations per grid cell per year.
We create the function 'mean_obs' to calculate this.

```{r}
# Function to calculate the mean of observations per grid cell per year
mean_obs <- function(data) {
  obs <- x <- NULL

  data %>%
    dplyr::mutate(x = mean(obs), .by = "cellCode") %>%
    dplyr::summarise(diversity_val = mean(x), .by = "year") %>%
    as.data.frame()
}
```

We get the following results:

```{r, message=FALSE}
mean_obs(processed_cube$data)
```

These values don’t tell us anything about reliability on their own.
Therefore, we perform a leave-one-species-out cross-validation (LOSO-CV) to investigate the impact of individual species being included in the calculation.

### Leave-one-species-out cross-validation

We use the `cross_validate_cube()` function to do this.
It relies on the following arguments:

-   **`data_cube`**:\
    The input data, either a processed data cube (from `b3gbi::process_cube()`), or just the dataframe inside (i.e. `processed_cube$data`).
    For faster computation, passing just the dataframe is recommended.

-   **`fun`**:\
    A user-defined function that computes the statistic(s) of interest from `data_cube`.
    This function should return a dataframe that includes a column named `diversity_val`, containing the statistic to evaluate.

-   **`grouping_var`**:\
    The column(s) used for grouping the output of `fun()`.
    For example, if `fun()` returns one value per year, use `grouping_var = "year"`.

-   **`out_var`**:\
    The variable used to leave out one group at a time during cross-validation.
    The default is `"taxonKey"` (typically representing species), which results in leave-one-species-out cross-validation.

-   **`progress`**:\
    Logical flag to show a progress bar.
    Set to `TRUE` to enable progress reporting; default is `FALSE`.

```{r}
cv_results <- cross_validate_cube(
  data_cube = processed_cube,
  fun = mean_obs,
  grouping_var = "year",
  out_var = "taxonKey"
)
```

```{r}
head(cv_results)
```

The RMSE is an average error measure we obtain for each year.
It remains very similar over the years.

```{r}
#| fig.alt: >
#|   RMSE for mean number of occurrences over time.
# Visualise mean errors
ggplot(cv_results, aes(x = as.factor(year))) +
  # Reference line
  geom_hline(yintercept = 0, colour = "black", linetype = "dashed") +
  # Plot RMSE
  geom_point(aes(y = rmse), colour = "red", size = 3) +
  # Settings
  labs(x = "", y = "RMSE") +
  theme_minimal()
```

Indeed, looking at the individual error values reveals similar patterns for every year.
The species with taxonkey 2481174 has an error of -2 in year 2011. This means that without this species the average number of occurrences per grid cell would be 2 untis lower than the estimate based on the full dataset (in this case around 46.8 instead of 48.8).

```{r}
#| fig.alt: >
#|   Species errors for mean number of occurrences over time.
# Get original estimates (based on full data cube)
original_estimates <- mean_obs(processed_cube$data)
original_estimates$label <- paste(
  "hat(theta) ==", round(original_estimates$diversity_val, 3)
)

# Visualise errors
ggplot(cv_results, aes(x = as.factor(year))) +
  # Reference line
  geom_hline(yintercept = 0, colour = "black", linetype = "dashed") +
  # Plot species keys
  geom_text(aes(y = error, colour = as.factor(taxonkey_out),
                label = as.factor(taxonkey_out))) +
  # Plot original estimates
  geom_label(data = original_estimates, aes(label = label),
             y = 0.9, parse = TRUE, size = 3) +
  # Settings
  scale_y_continuous(limits = c(NA, 1)) +
  labs(x = "", y = "LOSO-CV Error") +
  theme_minimal() +
  theme(legend.position = "")
```

There are two species that have a big effect on the calculation of our statistic.

-   2481174: [*Larus fuscus* Linnaeus, 1758](https://www.gbif.org/species/2481174)
-   2481139: [*Larus argentatus* Pontoppidan, 1763](https://www.gbif.org/species/2481139)

If we go back to the source of the data, we see that the data cube includes records from two large datasets strictly dedicated to these two species ([2481174](https://www.gbif.org/dataset/355b8ff9-7bd9-49c3-92af-f6741b8bd0cb), [2481139](https://www.gbif.org/dataset/6c860eb3-83ba-48c3-9328-a7b3c7a3c7b4)).

```{r}
index <- match("bird_cube_belgium_mgrs10", resources(b3data_package))
b3data_package$resources[[index]]$sources
```

This suggests that the two species are overrepresented in the data cube, likely due to targeted monitoring efforts.
Their disproportionate influence inflates indicator values, potentially distorting the true diversity signal.
Group-level sensitivity analysis like this helps uncover such sampling biases, ensuring that biodiversity indicators reflect ecological reality rather than arctifacts in the data.

## Advanced usage of `cross_validate_cube()`

### Cross-validate simple dataframes

As stated in the documentation, it is also possible to cross-validate over a dataframe.
In this case, set the argument `processed_cube = FALSE`.
This is implemented allow for flexible use of simple dataframes, while still encouraging the use of `b3gbi::process_cube()` as default functionality.

``` r
cv_results_df <- cross_validate_cube(
  data_cube = processed_cube$data,      # data.frame object
  fun = mean_obs,
  grouping_var = "year",
  out_var = "taxonKey",
  processed_cube = FALSE
)
```

### Optional arguments

The `cross_validate_cube()` function includes several optional parameters that give you more control over how the cross-validation is performed.
These arguments are not required for basic use, but can be helpful in specific scenarios.

-   **Cross-validation Method**: `crossv_method`\
    Determines how data is partitioned for cross-validation.

    -   `"loo"` (default):\
        Leave-one-out cross-validation.
        One category in `out_var` (e.g. one species) is excluded at a time.

    -   `"kfold"`:\
        K-fold cross-validation.
        All categories in `out_var` are split into `k` subsets, and each subset is excluded once while the others are used for analysis.\
        *Note: this method is experimental and results should be interpreted with caution.*

-   **Number of Folds**: `k`\
    Specifies the number of folds when using `"kfold"` cross-validation.\
    Default is `5`.
    Only used when `crossv_method = "kfold"`.

-   **Maximum Number of Categories**: `max_out_cats`\
    Sets an upper limit on the number of unique categories in `out_var` that will be excluded one-by-one during cross-validation.\
    Default is `1000`.
    This helps to prevent long runtimes for datasets with many unique categories.
    You can increase this if needed, but expect slower computation.
