---
title: "Visualising Spatial Trends"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This tutorial provides good practices regarding visualisation and interpretation of trends of indicators in space.
The methods discussed here are more broadly applicable, be for this tutorial we focus on occurrence cubes from which biodiversity indicators are derived.

## Calculating confidence intervals with dubicube

We reuse the example introduced in [bootstrap confidence interval calculation tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html) where we look at an occurrence cube of birds in Belgium between 2000 en 2024 using the MGRS grid at 10 km scale. We calculate confidence limits for the mean number of observations per grid cell over the years.

```{r, message=FALSE, warning=FALSE}
# Load packages
library(dubicube)

# Data loading and processing
library(frictionless) # Load example datasets
library(b3gbi)        # Process occurrence cubes

# General
library(ggplot2)      # Data visualisation
library(dplyr)        # Data wrangling
library(tidyr)        # Data wrangling
library(sf)           # Work with spatial objects
```

### Loading and processing the data

We load the bird cube data from the **b3data** data package using **frictionless** (see also [here](https://github.com/b-cubed-eu/b3data-scripts)).

```{r}
# Read data package
b3data_package <- read_package(
  "https://zenodo.org/records/15211029/files/datapackage.json"
)

# Load bird cube data
bird_cube_belgium <- read_resource(b3data_package, "bird_cube_belgium_mgrs10")
head(bird_cube_belgium)
```

We process the cube with **b3gbi**.
First, we select 3000 random rows to make the dataset smaller.
We only keep grid cells with more then 10 entries.
This is to reduce the computation time for this tutorial.

```{r}
set.seed(123)

# Make dataset smaller
rows <- sample(nrow(bird_cube_belgium), 3000)
bird_cube_belgium <- bird_cube_belgium[rows, ] %>%
  mutate(n_obs = n(), .by = "mgrscode") %>%
  filter(n_obs > 10) %>%
  select(-n_obs)

# Process cube
processed_cube <- process_cube(
  bird_cube_belgium,
  cols_occurrences = "n"
)
processed_cube
```

### Analysis of the data

Let's say we are interested in the mean number of observations per grid cell per year.
We create a function to calculate this.

```{r, echo=FALSE}
# nolint start: object_usage_linter.
```

```{r}
# Function to calculate statistic of interest
# Mean observations per grid cell
mean_obs_grid <- function(data) {
  data %>%
    dplyr::summarise(diversity_val = mean(obs), .by = "cellCode") %>%
    as.data.frame()
}
```

```{r, echo=FALSE}
# nolint end
```

We get the following results:

```{r}
mean_obs_grid(processed_cube$data)
```

On their own, these values don’t reveal how much uncertainty surrounds them. To better understand their variability, we use bootstrapping to estimate the distribution of the yearly means. From this, we can calculate bootstrap confidence intervals.

### Bootstrapping

We use the `bootstrap_cube()` function to perform bootstrapping (see also the [bootstrap tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).

```{r}
bootstrap_results <- bootstrap_cube(
  data_cube = processed_cube,
  fun = mean_obs_grid,
  grouping_var = "cellCode",
  samples = 1000,
  seed = 123
)
```

```{r}
head(bootstrap_results)
```

### Interval calculation

Now we can use the `calculate_bootstrap_ci()` function to calculate confidence limits (see also the [bootstrap confidence interval calculation tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).
We get a warning message for BCa calculation because we are using a relatively small dataset.

```{r}
ci_mean_obs <- calculate_bootstrap_ci(
  bootstrap_samples_df = bootstrap_results,
  grouping_var = "cellCode",
  type = c("perc", "bca", "norm", "basic"),
  conf = 0.95,
  data_cube = processed_cube,   # Required for BCa
  fun = mean_obs_grid                # Required for BCa
)

# Make interval type factor
ci_mean_obs <- ci_mean_obs %>%
  mutate(
    int_type = factor(
      int_type, levels = c("perc", "bca", "norm", "basic")
    )
  )
```
  
```{r}
head(ci_mean_obs)
```

## Visualising uncertainty in spatial trends

We can visualise the estimate and confidence levels in separate figures.

```{r}
# Read MGRS grid from repository
mgrs10_belgium <- st_read(
  "https://zenodo.org/records/15211029/files/mgrs10_refgrid_belgium.gpkg",
  quiet = TRUE
)
```

```{r}
ci_mean_obs %>%
  filter(int_type == "bca") %>%
  # Add MGRS grid
  left_join(mgrs10_belgium, by = join_by(cellCode == mgrscode)) %>%
  st_sf(sf_column_name = "geom", crs = st_crs(mgrs10_belgium)) %>%
  # Visualise result
  ggplot() +
  geom_sf(data = mgrs10_belgium) +
  geom_sf(aes(fill = est_original)) +
  # Settings
  labs(title = "Estimate", fill = "Legend") +
  theme_minimal()
```

```{r}
ci_mean_obs %>%
  filter(int_type == "bca") %>%
  # Add MGRS grid
  left_join(mgrs10_belgium, by = join_by(cellCode == mgrscode)) %>%
  st_sf(sf_column_name = "geom", crs = st_crs(mgrs10_belgium)) %>%
  # Visualise result
  ggplot() +
  geom_sf(data = mgrs10_belgium) +
  geom_sf(aes(fill = ul)) +
  # Settings
  labs(title = "Upper confidence limit", fill = "Legend") +
  theme_minimal()
```

```{r}
ci_mean_obs %>%
  filter(int_type == "bca") %>%
  # Add MGRS grid
  left_join(mgrs10_belgium, by = join_by(cellCode == mgrscode)) %>%
  st_sf(sf_column_name = "geom", crs = st_crs(mgrs10_belgium)) %>%
  # Visualise result
  ggplot() +
  geom_sf(data = mgrs10_belgium) +
  geom_sf(aes(fill = ll)) +
  # Settings
  labs(title = "Lower confidence limit", fill = "Legend") +
  theme_minimal()
```
