---
title: "Visualising Temporal Trends"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This tutorial provides good practices regarding visualisation and interpretation of trends of indicators over time.
The methods discussed here are more broadly applicable, be for this tutorial we focus on occurrence cubes from which biodiversity indicators are derived.
For visualisation and interpretation, we strongly rely on the functionality and concepts of the [**effectclass**](https://inbo.github.io/effectclass/articles/visualisation.html) package.

## Calculating confidence intervals with dubicube

We reuse the example introduced in [bootstrap confidence interval calculation tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html) where we calculate confidence limits for the mean number of observations per grid cell per year for birds in Belgium between 2000 en 2024 using the MGRS grid at 10 km scale.

```{r, message=FALSE, warning=FALSE}
# Load packages
library(dubicube)

# Data loading and processing
library(frictionless) # Load example datasets
library(b3gbi)        # Process occurrence cubes

# General
library(ggplot2)      # Data visualisation
library(dplyr)        # Data wrangling
library(tidyr)        # Data wrangling
```

### Loading and processing the data

We load the bird cube data from the **b3data** data package using **frictionless** (see also [here](https://github.com/b-cubed-eu/b3data-scripts)).

```{r}
# Read data package
b3data_package <- read_package(
  "https://zenodo.org/records/15211029/files/datapackage.json"
)

# Load bird cube data
bird_cube_belgium <- read_resource(b3data_package, "bird_cube_belgium_mgrs10")
head(bird_cube_belgium)
```

We process the cube with **b3gbi**.
First, we select 2000 random rows to make the dataset smaller.
This is to reduce the computation time for this tutorial.
We select the data from 2011 - 2020.

```{r}
set.seed(123)

# Make dataset smaller
rows <- sample(nrow(bird_cube_belgium), 2000)
bird_cube_belgium <- bird_cube_belgium[rows, ]

# Process cube
processed_cube <- process_cube(
  bird_cube_belgium,
  first_year = 2011,
  last_year = 2020,
  cols_occurrences = "n"
)
processed_cube
```

### Analysis of the data

Let's say we are interested in the mean number of observations per grid cell per year.
We create a function to calculate this.

```{r, echo=FALSE}
# nolint start: object_usage_linter.
```

```{r}
# Function to calculate statistic of interest
# Mean observations per grid cell per year
mean_obs <- function(data) {
  data %>%
    dplyr::mutate(x = mean(obs), .by = "cellCode") %>%
    dplyr::summarise(diversity_val = mean(x), .by = "year") %>%
    as.data.frame()
}
```

```{r, echo=FALSE}
# nolint end
```

We get the following results:

```{r}
mean_obs(processed_cube$data)
```

On their own, these values don’t reveal how much uncertainty surrounds them. To better understand their variability, we use bootstrapping to estimate the distribution of the yearly means. From this, we can calculate bootstrap confidence intervals.

### Bootstrapping

We use the `bootstrap_cube()` function to perform bootstrapping (see also the [bootstrap tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).

```{r}
bootstrap_results <- bootstrap_cube(
  data_cube = processed_cube,
  fun = mean_obs,
  grouping_var = "year",
  samples = 1000,
  seed = 123
)
```

```{r}
head(bootstrap_results)
```

### Interval calculation

Now we can use the `calculate_bootstrap_ci()` function to calculate confidence limits (see also the [bootstrap confidence interval calculation tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).
We get a warning message for BCa calculation because we are using a relatively small dataset.

```{r}
ci_mean_obs <- calculate_bootstrap_ci(
  bootstrap_samples_df = bootstrap_results,
  grouping_var = "year",
  type = c("perc", "bca", "norm", "basic"),
  conf = 0.95,
  data_cube = processed_cube,   # Required for BCa
  fun = mean_obs                # Required for BCa
)

# Make interval type factor
ci_mean_obs <- ci_mean_obs %>%
  mutate(
    int_type = factor(
      int_type, levels = c("perc", "bca", "norm", "basic")
    )
  )
```
  
```{r}
head(ci_mean_obs)
```

## Visualising temporal trends
### Error bars

The use of error bars is an easy and straightforward way to visualise uncertainty around indicator estimates.

```{r}
#| fig.alt: >
#|   Confidence intervals for mean number of occurrences over time.
ci_mean_obs %>%
  ggplot(aes(x = year)) +
  # Intervals
  geom_errorbar(aes(ymin = ll, ymax = ul),
                position = position_dodge(0.8), linewidth = 0.8) +
  # Estimates
  geom_point(aes(y = est_original),
             colour = "firebrick", size = 2) +
  # Settings
  labs(y = "Mean Number of Observations\nper Grid Cell") +
  scale_x_continuous(breaks = sort(unique(bootstrap_results$year))) +
  theme_minimal() +
  facet_wrap(~int_type)
```

However, the question remains which interval types should be calculated and/or reported.
A good idea is to compare different interval types next to each other together with the bootstrap distribution and the bootstrap bias (the difference between the estimate and the bootstrap estimate).

```{r}
#| fig.alt: >
#|   Confidence intervals with bootstrap distribution for mean number of occurrences over time.
# Get bias vales
bias_mean_obs <- bootstrap_results %>%
  distinct(year, estimate = est_original, `bootstrap estimate` = est_boot)

# Get estimate values
estimate_mean_obs <- bias_mean_obs %>%
  pivot_longer(cols = c("estimate", "bootstrap estimate"),
               names_to = "Legend", values_to = "value") %>%
  mutate(Legend = factor(Legend, levels = c("estimate", "bootstrap estimate"),
                         ordered = TRUE))
# Visualise
bootstrap_results %>%
  ggplot(aes(x = year)) +
  # Distribution
  geom_violin(aes(y = rep_boot, group = year),
              fill = alpha("cornflowerblue", 0.2)) +
  # Estimates and bias
  geom_point(data = estimate_mean_obs, aes(y = value, shape = Legend),
             colour = "firebrick", size = 2, alpha = 0.5) +
  # Intervals
  geom_errorbar(data = ci_mean_obs,
                aes(ymin = ll, ymax = ul, colour = int_type),
                position = position_dodge(0.8), linewidth = 0.6) +
  # Settings
  labs(y = "Mean Number of Observations\nper Grid Cell",
       x = "", shape = "Legend:", colour = "Interval type:") +
  scale_x_continuous(breaks = sort(unique(bootstrap_results$year))) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold"))
```

This informs us about the shape of the bootstrap distribution and the amount of bootstrap bias.
In combination with bootstrap interval theory, it can be decided which interval type(s) should be reported.
In general, because of the wide range of biodiversity indicator types, we recommend the use of percentile or BCa intervals, because they have no strong assumptions regarding the bootstrap distribution.
The BCa interval is recommended as it accounts for bias and skewness.
However, due to the jackknife estimation of the acceleration parameter, the calculation time is significantly longer.
The use of the normal and basic confidence intervals is not recommended, but could be used in combination with truncations or transformations.
The assumption of normality can be checked by making a Q-Q plot of the bootstrap replications (see further) (Davison & Hinkley, 1997).
An overview of the recommendations is provided in the table below.
This is not an exhaustive review of the topic, but based on existing literature and our preliminary results, these recommendations provide a useful starting point for selecting appropriate interval types.

<!-- spell-check: ignore:start -->
<table style="table-layout: auto; width: 100%;">
  <thead>
    <tr>
      <th style="white-space: nowrap;">Interval type</th>
      <th>Advantages</th>
      <th>Disadvantages</th>
      <th>References</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td><strong>Normal</strong></td>
      <td>
        <ul>
          <li>Simplicity</li>
          <li>Understanding of bootstrap and CI theory</li>
        </ul>
      </td>
      <td>
        <ul>
          <li>Assumes bootstrap distribution is normal</li>
          <li>Often, transformation needed for more accurate results</li>
          <li>Erratic coverage error in practice</li>
        </ul>
      </td>
      <td>
        Davison & Hinkley (1997, Ch. 5);<br>
        Efron & Tibshirani (1994, Ch. 13);<br>
        Hesterberg (2015)
      </td>
    </tr>
    <tr>
      <td><strong>Basic</strong></td>
      <td>
        <ul>
          <li>Simplicity</li>
          <li>Understanding of bootstrap and CI theory</li>
        </ul>
      </td>
      <td>
        <ul>
          <li>Assumes symmetric bootstrap distribution</li>
          <li>Typically substantial coverage error</li>
        </ul>
      </td>
      <td>
        Carpenter & Bithell (2000);<br>
        Davison & Hinkley (1997, Ch. 5);<br>
        Hesterberg (2015)
      </td>
    </tr>
    <tr>
      <td><strong>Percentile</strong></td>
      <td>
        <ul>
          <li>Simplicity</li>
          <li>No assumptions about bootstrap distribution</li>
          <li>Implicitly uses the existence of a good transformation</li>
        </ul>
      </td>
      <td>
        <ul>
          <li>Does not take bias into account</li>
          <li>Substantial coverage error if the distribution is not nearly symmetric</li>
        </ul>
      </td>
      <td>
        Carpenter & Bithell (2000);<br>
        Davison & Hinkley (1997, Ch. 5);<br>
        Efron & Tibshirani (1994, Ch. 13)
      </td>
    </tr>
    <tr>
      <td><strong>BCa</strong></td>
      <td>
        <ul>
          <li>No assumptions about bootstrap distribution</li>
          <li>Implicitly uses the existence of a good transformation</li>
          <li>Adjusts for bias</li>
          <li>Adjusts for skewness</li>
          <li>Smaller coverage error than the other methods</li>
        </ul>
      </td>
      <td>
        <ul>
          <li>Involved calculation of acceleration parameter <code>a</code></li>
          <li>Unstable coverage when sample size is small or <code>a</code> is small</li>
        </ul>
      </td>
      <td>
        Carpenter & Bithell (2000);<br>
        Davison & Hinkley (1997, Ch. 5);<br>
        Dixon (2001);<br>
        Efron & Tibshirani (1994, Ch. 14)
      </td>
    </tr>
  </tbody>
</table>
<!-- spell-check: ignore:end -->


### Smooth trends



### Fan plots

*check normality with qq plot*

## Visualising effects


## References


