---
title: "Whole-cube Bootstrap versus Group-specific Bootstrap"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

...

We reuse the example introduced in [bootstrap confidence interval calculation tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html) where we calculate confidence limits for the mean number of observations per grid cell per year for birds in Belgium between 2011 en 2020 using the MGRS grid at 10 km scale.

```{r, message=FALSE, warning=FALSE}
# Load packages
library(dubicube)

# Data loading and processing
library(frictionless) # Load example datasets
library(b3gbi)        # Process occurrence cubes

# General
library(ggplot2)      # Data visualisation
library(dplyr)        # Data wrangling
library(tidyr)        # Data wrangling
```

## Example data
### Loading and processing the data

We load the bird cube data from the **b3data** data package using **frictionless** (see also [here](https://github.com/b-cubed-eu/b3data-scripts)).

```{r}
# Read data package
b3data_package <- read_package(
  "https://zenodo.org/records/15211029/files/datapackage.json"
)

# Load bird cube data
bird_cube_belgium <- read_resource(b3data_package, "bird_cube_belgium_mgrs10")
head(bird_cube_belgium)
```

We process the cube with **b3gbi**.
First, we select 2000 random rows to make the dataset smaller.
This is to reduce the computation time for this tutorial.
We select the data from 2011 - 2020.

```{r}
set.seed(123)

# Make dataset smaller
rows <- sample(nrow(bird_cube_belgium), 2000)
bird_cube_belgium <- bird_cube_belgium[rows, ]

# Process cube
processed_cube <- process_cube(
  bird_cube_belgium,
  first_year = 2011,
  last_year = 2020,
  cols_occurrences = "n"
)
processed_cube
```

### Analysis of the data

Let's say we are interested in the mean number of observations per grid cell per year.
We create a function to calculate this.

```{r, echo=FALSE}
# nolint start: object_usage_linter.
```

```{r}
# Function to calculate statistic of interest
# Mean observations per grid cell per year
mean_obs <- function(data) {
  data %>%
    dplyr::mutate(x = mean(obs), .by = "cellCode") %>%
    dplyr::summarise(diversity_val = mean(x), .by = "year") %>%
    as.data.frame()
}
```

```{r, echo=FALSE}
# nolint end
```

We get the following results:

```{r}
mean_obs(processed_cube$data)
```

On their own, these values don’t reveal how much uncertainty surrounds them. To better understand their variability, we use bootstrapping to estimate the distribution of the yearly means. From this, we can calculate bootstrap confidence intervals.

## Whole-cube bootstrap
### Bootstrapping

We use the `bootstrap_cube()` function to perform bootstrapping (see also the [bootstrap tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).

```{r}
bootstrap_results <- bootstrap_cube(
  data_cube = processed_cube,
  fun = mean_obs,
  grouping_var = "year",
  samples = 1000,
  seed = 123
)
```

```{r}
head(bootstrap_results)
```

### Interval calculation

Now we can use the `calculate_bootstrap_ci()` function to calculate confidence limits (see also the [bootstrap confidence interval calculation tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).
We get a warning message for BCa calculation because we are using a relatively small dataset.

```{r}
ci_mean_obs <- calculate_bootstrap_ci(
  bootstrap_samples_df = bootstrap_results,
  grouping_var = "year",
  type = c("perc", "bca", "norm", "basic"),
  conf = 0.95,
  data_cube = processed_cube,   # Required for BCa
  fun = mean_obs                # Required for BCa
)

# Make interval type factor
ci_mean_obs <- ci_mean_obs %>%
  mutate(
    int_type = factor(
      int_type, levels = c("perc", "bca", "norm", "basic")
    )
  ) %>%
  mutate(method = "whole-cube")
```
  
```{r}
head(ci_mean_obs)
```

## Group-specific bootstrap
### Bootstrapping

We use the `bootstrap_cube()` function to perform bootstrapping (see also the [bootstrap tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).

```{r}
bootstrap_results_group <-processed_cube$data %>%
  split(processed_cube$data$year) %>%
  lapply(function(cube) {
    bootstrap_results <- bootstrap_cube(
      data_cube = cube,
      fun = mean_obs,
      grouping_var = "year",
      samples = 1000,
      seed = 123,
      processed_cube = FALSE
    )
    list(bootstrap_results = bootstrap_results, data = cube)
  })
```

### Interval calculation

Now we can use the `calculate_bootstrap_ci()` function to calculate confidence limits (see also the [bootstrap confidence interval calculation tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).
We get a warning message for BCa calculation because we are using a relatively small dataset.

```{r}
years <- unique(bootstrap_results_group$year)
ci_mean_obs_group_list <- vector(mode = "list", length = length(years))
for (i in seq_along(years)) {
  y <- years[i]

  ci_mean_obs_group_list[[i]] <- calculate_bootstrap_ci(
    bootstrap_samples_df = bootstrap_results_group %>%
      filter(year == y),
    grouping_var = "year",
    type = c("perc", "bca", "norm", "basic"),
    conf = 0.95,
    data_cube = processed_cube$data %>% # Required for BCa
      filter(year == y),   
    fun = mean_obs                      # Required for BCa
  )
}

ci_mean_obs_group_list <- bootstrap_results_group %>%
  lapply(
    calculate_bootstrap_ci,
    qgrouping_var = "year",
    type = c("perc", "bca", "norm", "basic"),
    conf = 0.95,
    data_cube = processed_cube$data %>% # Required for BCa
      filter(year == y),   
    fun = mean_obs)

# Make interval type factor
ci_mean_obs_group <- bind_rows(ci_mean_obs_group_list) %>%
  mutate(
    int_type = factor(
      int_type, levels = c("perc", "bca", "norm", "basic")
    )
  ) %>%
  mutate(method = "group-specific")
```
  
```{r}
head(ci_mean_obs_group)
```

## Comparsion

```{r}
# Visualise
bind_rows(
  bootstrap_results %>%
    mutate(method = "whole-cube"),
  bootstrap_results_group %>%
    mutate(method = "group-specific")
) %>%
  ggplot(aes(x = method, y = rep_boot)) +
  # Distribution
  geom_violin(fill = alpha("cornflowerblue", 0.2)) +
  # Settings
  facet_wrap(~year, ncol = 2, scales = "free") +
  labs(y = "Mean Number of Observations\nper Grid Cell") +
  theme_minimal()
```

```{r}
# Visualise
bind_rows(
  ci_mean_obs,
  ci_mean_obs_group
) %>%
  ggplot(aes(x = method, y = est_original)) +
  geom_point() +
  # Intervals
  geom_errorbar(
    aes(ymin = ll, ymax = ul, colour = int_type),
    position = position_dodge(0.8), linewidth = 0.8
  ) +
  # Settings
  facet_wrap(~year, ncol = 2, scales = "free") +
  labs(y = "Mean Number of Observations\nper Grid Cell") +
  theme_minimal()
```

## Whole-cube bootstrap2

Now we look at a function that does not depend on data from other years.

```{r, echo=FALSE}
# nolint start: object_usage_linter.
```

```{r}
# Function to calculate statistic of interest
# Mean observations per grid cell per year
mean_obs <- function(data) {
  data %>%
    group_by(year, cellCode) %>%
    dplyr::mutate(x = mean(obs)) %>%
    ungroup() %>%
    dplyr::summarise(diversity_val = mean(x), .by = "year") %>%
    as.data.frame()
}
```

```{r, echo=FALSE}
# nolint end
```

We get the following results:

```{r}
mean_obs(processed_cube$data)
```

### Bootstrapping

We use the `bootstrap_cube()` function to perform bootstrapping (see also the [bootstrap tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).

```{r}
bootstrap_results <- bootstrap_cube(
  data_cube = processed_cube,
  fun = mean_obs,
  grouping_var = "year",
  samples = 1000,
  seed = 123
)
```

```{r}
head(bootstrap_results)
```

### Interval calculation

Now we can use the `calculate_bootstrap_ci()` function to calculate confidence limits (see also the [bootstrap confidence interval calculation tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).
We get a warning message for BCa calculation because we are using a relatively small dataset.

```{r}
ci_mean_obs <- calculate_bootstrap_ci(
  bootstrap_samples_df = bootstrap_results,
  grouping_var = "year",
  type = c("perc", "bca", "norm", "basic"),
  conf = 0.95,
  data_cube = processed_cube,   # Required for BCa
  fun = mean_obs                # Required for BCa
)

# Make interval type factor
ci_mean_obs <- ci_mean_obs %>%
  mutate(
    int_type = factor(
      int_type, levels = c("perc", "bca", "norm", "basic")
    )
  ) %>%
  mutate(method = "whole-cube")
```
  
```{r}
head(ci_mean_obs)
```

## Group-specific bootstrap2
### Bootstrapping

We use the `bootstrap_cube()` function to perform bootstrapping (see also the [bootstrap tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).

```{r}
bootstrap_results_group <- processed_cube$data %>%
  split(processed_cube$data$year) %>%
  lapply(function(cube) {
    bootstrap_results <- bootstrap_cube(
      data_cube = cube,
      fun = mean_obs,
      grouping_var = "year",
      samples = 1000,
      seed = 123,
      processed_cube = FALSE
    )
    list(bootstrap_results = bootstrap_results, data = cube)
  })
```

### Interval calculation

Now we can use the `calculate_bootstrap_ci()` function to calculate confidence limits (see also the [bootstrap confidence interval calculation tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).
We get a warning message for BCa calculation because we are using a relatively small dataset.

```{r}
ci_mean_obs_group_list <- bootstrap_results_group %>%
  lapply(function(list) {
    calculate_bootstrap_ci(
      list$bootstrap_results,
      grouping_var = "year",
      type = c("perc", "bca", "norm", "basic"),
      conf = 0.95,
      data_cube = list$data,
      fun = mean_obs
    )
  })
    

# Make interval type factor
ci_mean_obs_group <- bind_rows(ci_mean_obs_group_list) %>%
  mutate(
    int_type = factor(
      int_type, levels = c("perc", "bca", "norm", "basic")
    )
  ) %>%
  mutate(method = "group-specific")
```
  
```{r}
head(ci_mean_obs_group)
```

## Comparsion2

```{r}
# Visualise
bind_rows(
  bootstrap_results %>%
    mutate(method = "whole-cube"),
  lapply(bootstrap_results_group, function(i) i$bootstrap_results) %>%
    bind_rows() %>%
    mutate(method = "group-specific")
) %>%
  ggplot(aes(x = method, y = rep_boot)) +
  # Distribution
  geom_violin(fill = alpha("cornflowerblue", 0.2)) +
  # Settings
  facet_wrap(~year, ncol = 2, scales = "free") +
  labs(y = "Mean Number of Observations\nper Grid Cell") +
  theme_minimal()
```

```{r}
# Visualise
bind_rows(
  ci_mean_obs,
  ci_mean_obs_group
) %>%
  ggplot(aes(x = method, y = est_original)) +
  geom_point() +
  # Intervals
  geom_errorbar(
    aes(ymin = ll, ymax = ul, colour = int_type),
    position = position_dodge(0.8), linewidth = 0.8
  ) +
  # Settings
  facet_wrap(~year, ncol = 2, scales = "free") +
  labs(y = "Mean Number of Observations\nper Grid Cell") +
  theme_minimal()
```

