---
title: "Whole-cube Bootstrap versus Group-specific Bootstrap"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

When calculating indicators from a biodiversity cube, we often want confidence intervals (CIs) using bootstrapping. However, depending on the indicator, we need to consider how to resample rows:

- Some indicators are per group (e.g., species, year, habitat), and each group is analysed independently.

- Other indicators require data from multiple groups simultaneously (e.g., community richness, turnover, or multi-species metrics).

The choice of bootstrap method affects both the accuracy and stability of confidence intervals. We distinguish whole-cube bootstrapping and group-specific bootstrapping.

### Whole-cube bootstrap

**Definition:** Resample all rows in the cube, regardless of species, year, or other grouping.

**Advantages:**

* Preserves correlations between groups (e.g., species co-occurrence, temporal dependencies).
* Appropriate for indicators that depend on multiple groups together (community-level metrics, multi-species diversity).

**Disadvantages:**

* Rare groups may end up with zero rows in some bootstrap replicates, leading to wider or undefined CIs.
* Variance for small groups may be inflated.

**Use case examples:**

* Community richness per site or habitat.
* Multi-species indicators (e.g., average occupancy across species).
* Temporal turnover indicators that rely on multiple years.

**Implementation in `dubicube`**

* Default use like vignettes and examples.

### Group-specific bootstrap

**Definition:** Subset the cube by the group of interest (e.g., species or year), then resample rows only within that group.

**Advantages:**

* Guarantees each replicate has rows for the group → stable CIs.
* Reflects within-group variability only.

**Disadvantages:**

* Ignores correlations with other groups.
* Variance may be slightly underestimated if the group’s presence is correlated with other groups.

**Use case examples:**

* Species-specific occupancy or habitat preference metrics.
* Year-specific indicators (e.g., annual richness).
* Small or rare groups where zero-row replicates would be problematic.

**Implementation in `dubicube`**

* Perform bootstrapping and interval calculation per group (e.g. using a for loop or `lapply()`).
* See further.

## An example of a group-specific analysis

We reuse the example introduced in [bootstrap confidence interval calculation tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html) where we calculate confidence limits for the mean number of observations per grid cell per year for birds in Belgium between 2011 en 2020 using the MGRS grid at 10 km scale.

```{r, message=FALSE, warning=FALSE}
# Load packages
library(dubicube)

# Data loading and processing
library(frictionless) # Load example datasets
library(b3gbi)        # Process occurrence cubes

# General
library(ggplot2)      # Data visualisation
library(dplyr)        # Data wrangling
library(tidyr)        # Data wrangling
```

### Loading and processing the data

We load the bird cube data from the **b3data** data package using **frictionless** (see also [here](https://github.com/b-cubed-eu/b3data-scripts)).

```{r}
# Read data package
b3data_package <- read_package(
  "https://zenodo.org/records/15211029/files/datapackage.json"
)

# Load bird cube data
bird_cube_belgium <- read_resource(b3data_package, "bird_cube_belgium_mgrs10")
head(bird_cube_belgium)
```

We process the cube with **b3gbi**.
First, we select 2000 random rows to make the dataset smaller.
This is to reduce the computation time for this tutorial.
We select the data from 2011 - 2020.

```{r}
set.seed(123)

# Make dataset smaller
rows <- sample(nrow(bird_cube_belgium), 2000)
bird_cube_belgium <- bird_cube_belgium[rows, ]

# Process cube
processed_cube <- process_cube(
  bird_cube_belgium,
  first_year = 2011,
  last_year = 2020,
  cols_occurrences = "n"
)
processed_cube
```

### Analysis of the data

Let's say we are interested in the mean number of observations per grid cell per year.
We create a function to calculate this.

> In contrast to the other vignettes, we now calculate this exclusively per year. Therefore, the group-specific bootstrap is more appropriate here.

```{r, echo=FALSE}
# nolint start: object_usage_linter.
```

```{r}
# Function to calculate statistic of interest
# Mean observations per grid cell per year
mean_obs <- function(data) {
  data %>%
    group_by(year, cellCode) %>%
    dplyr::mutate(x = mean(obs)) %>%
    ungroup() %>%
    dplyr::summarise(diversity_val = mean(x), .by = "year") %>%
    as.data.frame()
}
```

```{r, echo=FALSE}
# nolint end
```

We get the following results:

```{r}
mean_obs(processed_cube$data)
```

### Group-specific bootstrapping

We use the `bootstrap_cube()` function to perform bootstrapping (see also the [bootstrap tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).
However, we now split the data per year and perform bootstrapping per year using `lapply()`.
We store the data also in the resulting list to be used for interval calculation further on.

```{r}
bootstrap_results_group <- processed_cube$data %>%
  split(processed_cube$data$year) %>%
  lapply(function(cube) {
    bootstrap_results <- bootstrap_cube(
      data_cube = cube,
      fun = mean_obs,
      grouping_var = "year",
      samples = 1000,
      seed = 123,
      processed_cube = FALSE
    )
    list(bootstrap_results = bootstrap_results, data = cube)
  })
```

### Group-specific interval calculation

Now we can use the `calculate_bootstrap_ci()` function to calculate confidence limits (see also the [bootstrap confidence interval calculation tutorial](https://b-cubed-eu.github.io/dubicube/articles/bootstrap-method-cubes.html)).
Because BCa interval calculation relies on jackknifing, we also need to do this per group using `lapply()`.

```{r}
ci_mean_obs_group_list <- bootstrap_results_group %>%
  lapply(function(list) {
    calculate_bootstrap_ci(
      list$bootstrap_results,
      grouping_var = "year",
      type = c("perc", "bca", "norm", "basic"),
      conf = 0.95,
      data_cube = list$data,
      fun = mean_obs
    )
  })

# Make interval type factor
ci_mean_obs_group <- bind_rows(ci_mean_obs_group_list) %>%
  mutate(
    int_type = factor(
      int_type, levels = c("perc", "bca", "norm", "basic")
    )
  )
```
  
```{r}
head(ci_mean_obs_group)
```

We visualise the results.

```{r}
ci_mean_obs_group %>%
  ggplot(aes(x = year, y = est_original)) +
  # Intervals
  geom_errorbar(aes(ymin = ll, ymax = ul),
                position = position_dodge(0.8), linewidth = 0.8) +
  # Estimates
  geom_point(colour = "firebrick", size = 2) +
  # Settings
  labs(y = "Mean Number of Observations\nper Grid Cell") +
  scale_x_continuous(breaks = sort(unique(bootstrap_results$year))) +
  theme_minimal() +
  facet_wrap(~int_type)
```
